\section{Expanded Future Work and Next Steps}

Building on the initial research roadmap, this section provides a more detailed view of our future work, incorporating the insights from our competitive analysis, world model research, and future predictions. Our work is organized into three parallel tracks: \textit{Core Technology}, \textit{Platform and Product}, and \textit{Go-to-Market}.

\subsection{Core Technology Track}

This track is focused on advancing the state of the art in the core technologies that underpin our platform.

\begin{itemize}
    \item \textbf{GNN Research:} We will continue to push the boundaries of GNN scalability and expressiveness. Key research areas include distributed training for continent-scale graphs, techniques for mitigating over-smoothing and over-squashing, and architectures for handling heterophilic graphs.
    \item \textbf{World Model Development:} We will build on our initial research into world models, with a focus on adapting them to the unique characteristics of infrastructure networks. This will involve developing new representations for graph-structured data and incorporating domain-specific physics into the simulation.
    \item \textbf{Agent Architecture:} We will continue to refine our agent architecture, with a focus on improving long-horizon planning, multi-agent coordination, and safe decision-making under uncertainty. This will involve exploring techniques from hierarchical reinforcement learning, game theory, and causal inference.
\end{itemize}

\subsection{Platform and Product Track}

This track is focused on translating our core technology into a robust, scalable, and user-friendly platform.

\begin{itemize}
    \item \textbf{Data Platform:} We will continue to invest in our data platform, with a focus on building out our data ingestion, validation, and feature engineering pipelines. We will also develop tools for data provenance and governance, to ensure the quality and integrity of our data.
    \item \textbf{MCP and Tooling:} We will expand our library of MCP tools, providing our agents with a richer set of capabilities for interacting with the world. We will also develop a developer SDK to allow third parties to build their own tools and integrations.
    \item \textbf{User Experience:} We will work closely with our customers to design and build an intuitive and powerful user experience. This will include a web-based interface for visualizing data, configuring agents, and monitoring performance, as well as a rich API for programmatic access.
\end{itemize}

\subsection{Go-to-Market Track}

This track is focused on bringing our product to market and establishing AtlasPro AI as the category leader.

\begin{itemize}
    \item \textbf{Customer Partnerships:} We will continue to build deep partnerships with our initial customers, working closely with them to solve their most pressing problems and ensure their success.
    \item \textbf{Vertical Expansion:} While our initial focus is on the telecommunications and utilities sectors, we plan to expand into other verticals over time, including smart cities, transportation, and logistics.
    \item \textbf{Thought Leadership:} We will continue to publish our research in top-tier venues and engage with the broader research community. We believe that by sharing our work, we can accelerate the development of the field and establish AtlasPro AI as a thought leader in spatial intelligence.
\end{itemize}

This expanded roadmap reflects our ambition to not only build a successful company, but also to make a lasting contribution to the field of artificial intelligence. We are confident that by executing on this plan, we can achieve our goal of building an autonomous layer for the physical world.
